Long--term perspectives of social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 10, 2017, Umeå}

### 2017-06-07 Umea.Rnw
Revised slides

\title{Social inequality in mortality among adults and elderly in
  northern  Sweden 1851--2013\footnote{Seminar June 10, 2017, ICASS, Umeå}
}

@@

\author{Sören Edvinsson and Göran Broström}

@@

\begin{frame}{Questions}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased in the adult and elderly
  population?

@@

\item Are there any \emp{gender differences} in the effect of social position?

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively?
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?

\end{enumerate}

@@

\begin{frame}{Top incomes}

@@

\item \emp{Data} management
\item \emp{Modeling} considerations
\item \emp{Analysis}
\item \emp{Results}
\item \emp{Alternative} Results
\end{enumerate}

@@

\begin{frame}{Here!}

@@

Skellefteå:

\item Small town, heavy industry in the nineteenth century.

@@

\item Not so small town, older. Center for administration, education, and military.

@@

Rural areas:

\item Very similar: Small farms dominating.

@@

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[\sc Hisclass] elite, middle, worker.
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, seven categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{frame}{\sc Hisclass}

@@

\begin{frame}{No proportional hazards (wrt \sc Hisclass)!}

\begin{itemize}
\item Fit each {\sc Hisclass} \emp{separately}, for each time period.
\item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
\end{itemize}

@@

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
  \item \emp{\tt hisclass}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}
\end{frame}

@@

  %\begin{center}
\includegraphics[height = 7.5cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm]{figs/slide-cumswold-1}
   %\end{center}
\end{frame}

@@

\begin{frame}{Conclusion}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item There are \emp{gender differences} in the social pattern
\item No difference in the social pattern in the two studied age groups.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item \emp{Education} and \emp{income} are the \emp{most important}
  variables during late
  time period.
\end{itemize}

@@

>>>>

\title{Long--term perspectives of social inequality in mortality among adults and elderly in
  northern  Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 10, 2017, Umeå}
}

@@

\author{Sören Edvinsson and Göran Broström \\
Centre for Ageing and Demographic Research, Umeå University}

@@

\begin{frame}{Aims}

To investigate the association between social class and mortality among adults and elderly from the 19th century to the present and how it is connected to socio-economic inequality in society.
This is analysed according to 
\begin{itemize}
\item gender
\item age group (working age vs retired)
\item cause of death (cardiovascular, cancer)
\item (education 1990--2013)
\item (income 1986-2005)

% \item Has \emp{inequality in mortality} between \emp{social classes}
%   increased in the adult and elderly
%   population?

@@

% \item Are there any \emp{gender differences} in the effect of social position?
% 
% \item Is social position \emp{equally important} among the \emp{retired} population as
%   for those in \emp{working age}?
% \item Is there a difference between classes in mortality from
%   \emp{cardiovascular} diseases and \emp{cancers} respectively?
% \item Are there large differences in survival depending on level of
%   \emp{education} and \emp{income} categories?

@@

\begin{frame}{Main results}

\begin{itemize}
\item A consistent mortality difference according to social class appears surprisingly late in our regions and has no clear connection to the development of income inequality.
\item However, there is a strong gender difference in the development of social inequality.
\item We argue that health-related behavior is an important aspect in historical as well as present-day mortality, and that such behaviours are based on strong gender and class expectations.
\end{itemize}

\begin{frame}{Economic inequality}

@@

\item \emp{Data} management
\item \emp{Modeling} considerations
\item \emp{Analysis}
\item \emp{Results}
\item \emp{Implications} of the results
\end{enumerate}

@@

\begin{frame}{Two regions in Västerbotten}

@@

Remote area in northern Sweden, dominated by agriculture until well into the 20th century. 19th century communications were rudimentary.

Skellefteå:
\begin{itemize}

\item Small town, industrialization of the region during the 20th century.

@@

\item Center for administration, education, and military. University from 1965

@@

% Rural areas:
% 
% \begin{itemize}
% 
% \item Very similar: Small farms dominating.
% 
% \end{itemize}

@@

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[Social class] elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, three categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{frame}{\sc Social class}

@@

\begin{frame}{No proportional hazards (wrt class)!}

\begin{itemize}
\item Fit each social class \emp{separately}, for each time period.
\item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
\end{itemize}

@@

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
  \item \emp{\tt class}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}
\end{frame}

@@

  %\begin{center}
\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswold-1}
   %\end{center}
\end{frame}

@@

\begin{frame}{Summarizing the results}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
\item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection between income inequality and social differences in mortality.
\end{itemize}

@@

\begin{frame}{Some implications}

\begin{itemize}
\item The development of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
\item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
\item The role of \emp{living standard} for mortality decline (McKeown).
\item \emp{Gender} and mortality (Watkins, Mckinnon, Janssens).
\item Social conditions as \emp{fundamental causes of death} (Link/Phelan).

%\begin{frame}{Conclusion}

% \begin{itemize}
% \item \emp{Improved} survival characterised all {\sc Hisclass}es.
% \item There are \emp{gender differences} in the social pattern
% \item No difference in the social pattern in the two studied age groups.
% \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
% \item \emp{Educational} and \emp{income} income differences during the last decades conform with what we found for class differences.
% \end{itemize}

%\end{frame}
